---
title: "Exploration and analysis"
author: "Anne Wilson"
date: "2024-02-29"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
knit: (function(input_file, encoding) {
    out_dir <- 'docs';
    rmarkdown::render(input_file,
      encoding=encoding,
      output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
```

```{r echo = FALSE, include = FALSE}
library(tidyverse)
bike_share_2023_processed <- read_rds("bike_share_2023_processed.RDS")
```
## Analysis
First I calculate the total number of rides taken by each of the two groups, members and casual riders.
```{r}
bike_share_2023_processed %>% 
  group_by(member_casual) %>% 
  summarise(number_of_rides = n())
```

### Ride duration
Now I look to see the range of ride lengths in each of the two groups, members and casual riders. 
```{r}
bike_share_2023_processed %>% 
  group_by(member_casual) %>% 
  summarise(min_duration = min(ride_length), max_duration = max(ride_length))
```

Note that the min-duration of 2 minutes here is artificial, as rides of length less than two minutes were removed when the data was cleaned as there were a number of extremely short rides that could not have been real rides.

Histograms are generated to show the number of rides of different durations for members and for casual riders. The rides have been split into three groups - those less than an hour long, those between one and twelve hours, and those that last over twelve hours.
```{r}
bike_share_2023_processed %>% 
  filter(ride_length < 60) %>% 
  ggplot(mapping = aes(x = as.numeric(ride_length), fill = member_casual)) +
  geom_histogram(binwidth = 5, boundary = 0) +
  facet_wrap(~member_casual) +
  labs(title = "Ride Length",subtitle = "Ride length less than one hour.", x = "Ride length (minutes)", y = "Number of rides")
```

```{r}
bike_share_2023_processed %>%
  filter(ride_length > 60 & ride_length < 720) %>% 
  ggplot(mapping = aes(x = as.numeric(ride_length)/60, fill = member_casual)) +
  scale_x_continuous(breaks = c(2,4,6,8,10,12), labels=c("2", "4", "6", "8", "10", "12")) +
  coord_cartesian(xlim = c(1, 12))  +
  geom_histogram(binwidth = 1, boundary = 1) +
  facet_wrap(~member_casual) +
  labs(title = "Ride Length",subtitle = "Ride length between 1 and 12 hours.", x = "Ride length (hours)", y = "Number of rides")
```

```{r}
bike_share_2023_processed %>% 
  filter(ride_length > 720) %>% 
  ggplot(mapping = aes(x = as.numeric(ride_length)/60, fill = member_casual)) +
  geom_histogram(binwidth = 12, boundary = 12) +
  coord_cartesian(xlim = c(12, 204)) +
  scale_x_continuous(breaks = c(24,48,72,96,120,144,168,192)) +
  facet_wrap(~member_casual) +
  labs(title = "Ride Length",subtitle = "Ride length greater than 12 hours.", x = "Ride length (hours)", y = "Number of rides")
```

These histograms show that the majority of rides are less than an hour long for both casual riders and members, but a higher proportion of the rides made by casual riders, rather than members, are longer than an hour. There are very few rides that are longer than 24 hours for both groups. It should be remembered that there are many more rides by members than by casual riders. 

Most of the rides are less than 15 minutes long, and I now calculate the percentage of these shorter rides for each group.
```{r}
casual_in_interval <- nrow(bike_share_2023_processed[(bike_share_2023_processed$member_casual == "casual") &
                                 (bike_share_2023_processed$ride_length < 15),])
total_casual <- nrow(bike_share_2023_processed[(bike_share_2023_processed$member_casual == "casual"),])

member_in_interval <- nrow(bike_share_2023_processed[(bike_share_2023_processed$member_casual == "member") &
                                 (bike_share_2023_processed$ride_length < 15),])
total_member <- nrow(bike_share_2023_processed[(bike_share_2023_processed$member_casual == "member"),])

percent_casual_in_interval <- 100 * casual_in_interval / total_casual
percent_member_in_interval <- 100 * member_in_interval / total_member

percent_casual_in_interval
percent_member_in_interval
```

From which, it can be seen that 74% of all rides by annual members are shorter than 15 minutes, whereas only 55% of rides by casual riders are that short.

### When the bikes are ridden
I now consider on what day of the week the bikes are ridden, again split by casual riders and members. This is done by looking at the day_of_week column which gives the day on which the bike is taken out. There are some rides that cross into the next day but for this analysis only the start day of the ride is considered.
```{r}
bike_share_2023_processed %>%
  ggplot(mapping = aes(x = day_of_week, fill = member_casual)) +
  geom_bar() +
  facet_wrap(~member_casual) +
  labs(title = "Number of Rides for each Day of the Week", x = "Day of week", y = "Number of rides (thousands)") +
  scale_x_continuous(breaks = c(1,2,3,4,5,6,7), labels=c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat")) +
  scale_y_continuous(breaks = c(0, 100000, 200000, 300000, 400000), labels = c("0", "100", "200", "300", "400"))
```

It can be seen the number of rides taken by casual users is highest at the weekend, whereas there are more rides by members during the week.

Next I do the same analysis but for the months to see if there is any difference in how members and casual riders use the bikes across the year.
```{r}
bike_share_2023_processed %>%
  ggplot(mapping = aes(x = month, fill = member_casual)) +
  geom_bar() +
  facet_wrap(~member_casual) +
  labs(title = "Number of Rides Across the Year", x = "Month", y = "Number of rides (thousands)") +
  scale_x_continuous(breaks = c(1,3,5,7,9,11), labels=c("Jan", "Mar", "May", "Jul", "Sep","Nov")) +
  scale_y_continuous(breaks = c(0, 100000, 200000, 300000, 400000), labels = c("0", "100", "200", "300", "400"))
```

The pattern of rides across the year is very similar for members and casual riders, with fewer rides by both groups in the winter months.

Finally I consider the time of day that the rides start.
```{r}
bike_share_2023_processed %>%
  ggplot(mapping = aes(x = hour(started_at), fill = member_casual)) +
  geom_bar() +
  facet_wrap(~member_casual) +
  labs(title = "Number of Rides by Time of Day", x = "Time", y = "Number of rides (thousands)") +
  scale_x_continuous(breaks = c(2,6,10,14,18,22), labels=c("2:00", "6:00", "10:00", "14:00", "18:00", "22:00")) +
  scale_y_continuous(breaks = c(0, 100000, 200000, 300000, 400000), labels = c("0", "100", "200", "300", "400"))
```

There are clear peaks at the start and end of the working day for members but not for casual riders, which would suggest that fewer casual riders are using the bikes to commute to work (or they are working more variable hours).

### Locations
It would be interesting to see if casual riders use different stations to members. To do this I create a new data frame (station_ride_nums) giving the number of rides starting at each station.
```{r}
station_ride_nums <- bike_share_2023_processed %>%
  select(start_station_id, start_station_name, start_lat, start_lng, member_casual) %>% 
  group_by(member_casual, start_station_id, start_station_name) %>%
  # Average the start_lat and start_lng because they differ in the later digits for the same station.
  summarise(number_of_rides = n(), start_latitude = round(mean(start_lat), 3), start_longitude = round(mean(start_lng), 3)) %>% 
  arrange(-number_of_rides)

head(station_ride_nums, 10)
```

I save this data frame to a csv file so that I can create a map visualisation in Tableau.
```{r}
write_csv(station_ride_nums, "station_ride_nums.csv")
write_rds(station_ride_nums, "station_ride_nums.RDS")
```

Next I look at the top starting stations for casual users and for members and what proportion these make up of all rides. A data frame is created with the numbers of rides for each station and the percentage of casual or member rides.
```{r}
df <- station_ride_nums %>% 
  group_by(member_casual) %>% 
  reframe(start_station_id, start_station_name, total_rides = sum(number_of_rides), number_of_rides, percentage_of_rides = (number_of_rides / total_rides) * 100)
```

The data frame is then filtered to show just members, and sorted by the number of rides.
```{r}
members <- df %>% 
  filter(member_casual == "member") %>% 
  arrange(-number_of_rides)
head(members, 30)
```

The same is then done for casual riders.
```{r}
casual_riders <- df %>% 
  filter(member_casual == "casual") %>% 
  arrange(-number_of_rides)
head(casual_riders, 30)
```

The most used station for casual riders is Streeter Dr & Grand Ave, which accounts for 2.8% of all rides made by casual users. In contrast, the most used station for members (Kingsbury St & Kinzie St) accounts for only 0.9% of rides by members.

The same analysis for end stations is carried out below.
```{r}
end_ride_nums <- bike_share_2023_processed %>%
  # Average the end_lat and end_lng because they differ in the later digits for the same station.
  select(end_station_id, end_station_name, end_lat, end_lng, member_casual) %>% 
  group_by(member_casual, end_station_id, end_station_name) %>%
  summarise(number_of_rides = n(), end_latitude = round(mean(end_lat), 3), end_longitude = round(mean(end_lng), 3))

df <- end_ride_nums %>% 
  group_by(member_casual) %>% 
  reframe(end_station_id, end_station_name, total_rides = sum(number_of_rides), number_of_rides, percentage_of_rides = (number_of_rides / total_rides) * 100)
```

Filter and sort for members.
```{r}
members <- df %>% 
  filter(member_casual == "member") %>% 
  arrange(-number_of_rides)
head(members, 30)
```

Filter and sort for casual riders.
```{r}
casual_riders <- df %>% 
  filter(member_casual == "casual") %>% 
  arrange(-number_of_rides)
head(casual_riders, 30)
```

The results look similar for the end stations, with the casual riders more concentrated on a few stations.

### Numbers of rides made on electric and classic bikes
It may be that casual riders make more trips on either the electric or the classic bikes than members do, so I plot the total number of rides on each sort of bike for members and casual riders.
```{r}
ggplot(bike_share_2023_processed, mapping = aes(x = member_casual, fill = rideable_type)) +
  geom_bar(position = position_dodge()) +
  labs(title = "Number of rides on classic and electric bikes", x = "", y = "Number of rides (millions)") +
  scale_y_continuous(breaks = c(0, 500000, 1000000, 1500000, 2000000), labels = c("0", "0.5", "1.0", "1.5", "2.0"))
```

It can be seen that the proportion rides made by casual riders using electric bikes is greater than that of members. The exact percentages can be calculated to make this clearer.

```{r}
electric_member_count <- 
  nrow(bike_share_2023_processed[(bike_share_2023_processed$member_casual == "member") & 
                                    (bike_share_2023_processed$rideable_type == "electric_bike"),])
classic_member_count <- 
  nrow(bike_share_2023_processed[(bike_share_2023_processed$member_casual == "member") & 
                                    (bike_share_2023_processed$rideable_type == "classic_bike"),])
electric_casual_count <- 
  nrow(bike_share_2023_processed[(bike_share_2023_processed$member_casual == "casual") & 
                                    (bike_share_2023_processed$rideable_type == "electric_bike"),])
classic_casual_count <- 
  nrow(bike_share_2023_processed[(bike_share_2023_processed$member_casual == "casual") & 
                                    (bike_share_2023_processed$rideable_type == "classic_bike"),])

percent_member_electric <- 100 * electric_member_count / (electric_member_count + classic_member_count)
percent_casual_electric <-  100 * electric_casual_count / (electric_casual_count + classic_casual_count)

percent_member_electric
percent_casual_electric
```

It can be seen that the difference is small, with 35% of rides by members being on electric bikes, and 38% of rides by casual riders.

Next the number of bike rides on each type of bike is plotted for each day of the week.
```{r}
ggplot(bike_share_2023_processed, mapping = aes(x = day_of_week, fill = rideable_type)) +
  geom_bar(position = position_dodge()) +
  facet_wrap(~ member_casual) +
  labs(title = "Rides on classic and electric bikes by day of week", x = "Day", y = "Number of rides (thousands)") +
  scale_x_continuous(breaks = c(1, 2, 3, 4, 5, 6, 7), labels = c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat")) +
  scale_y_continuous(breaks = c(0, 100000, 200000, 300000), labels = c("0", "100", "200", "300"))
```

The pattern of usage of electric bikes is broadly similar to that for classic bikes, but the increase in the number of rides at weekends seen for casual riders is larger for the classic bikes than for electric bikes. The graph suggests that more weekend riders like to ride the classic rides, whereas in the week, while classic bikes are still used more often, the difference is less pronounced. 
Now I have a look at how this varies across the year.
```{r}
ggplot(bike_share_2023_processed, mapping = aes(x = month, fill = rideable_type)) +
  geom_bar(position = position_dodge()) +
  facet_wrap(~ member_casual) +
  labs(title = "Rides on classic and electric bikes by month of year", x = "Month", y = "Number of rides (thousands)") +
  scale_x_continuous(breaks = c(1,3,5,7,9,11), labels=c("Jan", "Mar", "May", "Jul", "Sep","Nov")) +
  scale_y_continuous(breaks = c(0, 50000, 100000, 150000, 200000, 250000), labels = c("0", "50", "100", "150", "200", "250")) 
```

And finally by time of day.
```{r}
ggplot(bike_share_2023_processed, mapping = aes(x = hour(started_at), fill = rideable_type)) +
  geom_bar(position = position_dodge()) +
  facet_wrap(~ member_casual) +
  labs(title = "Rides on classic and electric bikes by time of day", x = "Time", y = "Number of rides (thousands)") +
  labs(title = "Number of Rides by Time of Day", x = "Time", y = "Number of rides (thousands)") +
  scale_x_continuous(breaks = c(2,6,10,14,18,22), labels=c("2:00", "6:00", "10:00", "14:00", "18:00", "22:00")) + 
  scale_y_continuous(breaks = c(0, 50000, 100000, 150000, 200000), labels = c("0", "50", "100", "150", "200"))
```

These two plots show that a larger proportion of electric bikes are used in the winter and this effect is more pronounced for the casual riders. Also, for both members and casual riders, a higher proportion of rides are made by electric bikes at night. Since there are far fewer rides made in the winter and at night, it is possible this change in the proportion of electric bikes used is partly driven by availability. If there is a limited number of electric bikes, it may be that during summer days they are all in use and only classic bikes are available for many users.

### Some summary statistics.
I calculate the mean ride length for members and for casual riders. I filter the data here to look at just those rides that are shorter than 24 hours to avoid the few exceptionally long rides skewing the data.
```{r}
bike_share_2023_processed %>% 
  filter(ride_length < 1440) %>% 
  group_by(member_casual) %>% 
  summarise(mean_ride_length = mean(ride_length))
```

And then look at the mean ride length for each day of the week for the two groups to see if there is a difference across the week. I filter the data here to look at just those rides that are shorter than 24 hours to avoid the few exceptionally long rides skewing the data.
```{r}
days_of_rides <- bike_share_2023_processed %>%
  filter(ride_length < 1440) %>% 
  group_by(day_of_week, member_casual) %>% 
  summarise(mean_ride_length = mean(ride_length))

ggplot(days_of_rides, mapping = aes(x = day_of_week, y = as.numeric(mean_ride_length), fill = member_casual)) +
  geom_col() +
  facet_wrap(~member_casual) +
  labs(title = "Mean ride duration for each day of the week", x = "Day of week", y = "Mean ride length (minutes)") +
  scale_x_continuous(breaks = c(1,2,3,4,5,6,7), labels=c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat"))
```

It can be seen that the rides are longer at the weekends for both members and casual riders, but for members the mean ride length is very similar for all week days whereas, for casual users, it is higher at the beginning and end of the week. 

Next I consider the mean ride lengths by time of day (with rides over 24 hours excluded).
```{r}
times_of_rides <- bike_share_2023_processed %>%
  filter(ride_length < 1440) %>% 
  group_by(hour_of_day = hour(started_at), member_casual) %>% 
  summarise(mean_ride_length = mean(ride_length))

ggplot(times_of_rides, mapping = aes(x = hour_of_day, y = as.numeric(mean_ride_length), fill = member_casual)) +
  geom_col() +
  facet_wrap(~member_casual) +
  labs(title = "Mean ride duration for each hour of the day", x = "Time of day", y = "Mean ride length (minutes)") +
  scale_x_continuous(breaks = c(2,6,10,14,18,22), labels=c("2:00", "6:00", "10:00", "14:00", "18:00", "22:00"))
```

There is much more variation across the day for the casual riders, with rides starting between 9am and 3pm having the longest average ride length, a peak which is not seen for the members.

### Conclusions
The data shows that there are more rides made by casual riders at weekends and during the middle of the day, and that those rides tend to be longer. In contrast rides made by members tend to be shorter and peak at the start and end of the working day. This information, taken together with the popularity of the stations at tourist destinations for casual riders, suggest that a larger proportion of the casual rides are for leisure, whereas the members' rides are more often short rides in the week, including commuting to and from work.

