---
title: "Processing the bike-share data"
author: "Anne Wilson"
date: "2024-02-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
knit: (function(input_file, encoding) {
    out_dir <- 'docs';
    rmarkdown::render(input_file,
      encoding=encoding,
      output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
```

## Loading required packages
```{r echo = FALSE}
library(tidyverse)
```

```{r}
bike_share_2023_cleaned <- read_rds("bike_share_2023_cleaned.RDS")
```

## Transforming the data
First some calculated columns are added to the data. These give the ride length, day of week the ride started and the month the ride started. The ride lengths are rounded to the nearest minute.
```{r}
bike_share_2023_processed <-  bike_share_2023_cleaned %>% 
  add_column(ride_length = round(difftime(bike_share_2023_cleaned$ended_at, bike_share_2023_cleaned$started_at, units = "mins"))) %>% 
  add_column(day_of_week = wday(bike_share_2023_cleaned$started_at)) %>% 
  add_column(month = month(bike_share_2023_cleaned$started_at))
```

Look at the data.
```{r}
glimpse(bike_share_2023_processed)
```
Now a summary of the ride lengths is generated to look at the mean, median, minimum and maximum.
```{r}
bike_share_2023_processed %>% 
  summarise(mean_ride_length = mean(ride_length), median_ride_length = median(ride_length), min_ride_length = min(ride_length), max_ride_length = max(ride_length))
```
There is at least one ride of length zero and there is a ride of length 12136 minutes, that is over 8 days - this might be an outlier and may cause problems with the analysis.
Rides of less than one minute are unlikely to be genuine rides. Filter to look at the rides that are less than one minute.
```{r}
bike_share_2023_processed %>% 
  filter(ride_length < 1)
```

There are 55907 of these rows with ride length less than one minute. These are deleted to avoid them skewing the final analysis.
```{r}
bike_share_2023_processed <- bike_share_2023_processed %>% 
  filter(ride_length > 1)
```

Next the data is filtered to see how many rides there are that are more than 24 hours (1440 minutes).
```{r}
bike_share_2023_processed %>% 
  arrange(-ride_length) %>% 
  filter(ride_length > 1440)
```

While there are not many rides longer than 24 hours, there are over 100 of them and they are likely to be real data, so these will be kept but they may need to be handled separately for some of the analysis.

Now the data is ready to be analysed, so the processed data in the data frame bike_share_2023_processed is saved to RDS and CSV files.
```{r}
write_rds(bike_share_2023_processed, file = "bike_share_2023_processed.RDS")
write_csv(bike_share_2023_processed, "bike_share_2023_processed.csv")
```



